\textit{Supporting Information for}: Geometric models reveal behavioral and neural signatures of transforming experiences into memories

### 2020-10-23 paper/supplementary_information.tex
recompiled diff file

@@

@@

\newpage
\begin{figure}[tp]
\centering
%\includegraphics[width=1\textwidth]{figs/}
\caption{\small \textbf{Methods detail for recall trajectory analysis displayed in Figure~\trajectories B} \textbf{A.} This panel replicates Figure~\trajectories B in the \textit{Main text}, but with two additions.  First, individual participants' recall trajectories are displayed (faintly) as light gray lines.  Second, three locations on the trajectory have been highlighted (blue, yellow, and red circles).  \textbf{B.}  These zoomed-in views of the locations highlighted in Panel A show the average trajectory (black) and individual participants' trajectories (gray lines) that intersect the given region of topic space.  \textbf{C.} For each circular region of topic space tiling the 2D embedding plane displayed in Panel A, we compute the distribution of angles formed between each participant's trajectory segment (i.e., the point at which the trajectory enters and exists the region of topic space) and the $x$-axis.  The distributions of angles for these three example regions are displayed in the colored rose plots.  We use Rayleigh tests to assign an arrow direction, length, and color for that region of topic space.  Arrows displayed in blue are significant at the $p < 0.05$ level (corrected).  The arrow directions are oriented according to the circular means of each distribution, and the arrow lengths are proportional to the lengths of those mean vectors.  The example regions have been oriented from left to right in decreasing order of consistency across participants.}
\label{fig:arrows-methods}
\end{figure}
\vspace*{\fill}

>>>>

\subsubsection*{Binary features}
Two categories of annotations, ``Music presence'' and ``Indoor vs outdoor,'' comprise a single word that can each take on one of two possible values (music: ``yes'' or ``no''; indoor vs outdoor: ``indoor'' or ``outdoor'').  Participants are unlikely to use the words ``yes'' or ``no'' to specifically refer to the presence or absence of music when they recount the episode.  In fact, the word ``no'' is filtered out at the start of our analysis pipeline (see \textit{Methods}), since it appears so frequently throughout the annotations and recalls that it cannot provide reliable event-specific information.

In contrast, the word ``yes'' is a comparatively rare word (both in the episode annotations and participants’ recall transcripts). The rarity of the word ``yes'' in the annotations makes it a potentially informative feature for the topic models.  For example, the presence of music often co-occurs with the presence of specific characters, actions, locations, and plot elements.  In this way, the ``Music presence'' annotations provide a (likely relatively weak) signal for when these associated themes appear.  Further, when participants reference those themes in their recalls through their uses of other words associated with those themes (e.g., even if they don’t specifically use the word ``yes''), our modeling framework will still ``match'' references to music-related themes (i.e., semantic features or topics in the episode that co-occur with the presence or absence of music) with other words associated with those semantic features or topics.

The ``Indoor vs outdoor'' annotations are treated similarly.  Again, although participant may not refer to a particular scene as having taken place indoors versus outdoors, they may refer to other themes that co-occur with these annotations.  In general, the notion that annotations and recalls do not need to use the same words (or in the same ways)—is a central feature of our modeling framework.  We think solving the ``matching problem'' (i.e., labeling specific things participants say with specific events they experienced) in a way that is robust to wording differences is an important advance in studying naturalistic memory behaviors.

@@

\subsection*{Additional comments on the searchlight analyses (Fig.~\brains)}
The final ``stage'' shown for the episode template pipeline (Fig. \brains A) displays the correlations between pairs of episode topic vectors at nearby timepoints, while the final stage shown for the recall template pipeline (Fig. \brains B) displays the correlations between pairs of recall topic vectors at nearby timepoints.  The ``proximal correlation matrices'' we mention in the figure and main text denote that we have taken temporal autocorrelation matrices and masked out everything above the $n$th diagonal (where $n$ is the duration, in TRs, of the longest episode event).  When we say ``correlations at nearby timepoints,'' we are referring to the unmasked parts of these proximal correlation matrices.

To compute proximal correlation matrices for the episode, or for a given searchlight’s activity patterns, we simply correlate the topic vectors (or voxel responses) from every pair of timepoints, and then mask out anything beyond the $n$th diagonal.  In this way, the RSA analysis that we designed to identify searchlights whose responses show a similar (proximal) correlation structure to the episode’s topics (Fig. \brains C) does not require any further temporal alignment, since both the topic timeseries and voxel responses are computed at the same timepoints.

However, computing the proximal correlation matrix for participants’ recalls requires an additional ``warping'' step to bring the behavioral data (during recall) and neural data (while viewing the episode) into temporal alignment.  For example, different participants take different amounts of time to recount the episode, and their transcripts have different lengths.  These differences occur at the ``episode'' level (i.e., with respect to total recall duration and/or transcript length) and also at the ``event'' level (i.e., a given participant's recounting of a particular event may be more or less detailed, and take more or less time, than another participant's recounting of the same event).  The purpose of the warping step is to temporally stretch or compress the topic timecourse of each participants’ recounting so that it is temporally aligned with the voxel responses recorded as the participants were watching the episode.  That warping step is what we are highlighting in the two rightmost matrices in Figure~\brains B.

The transition between the 1st and 2nd stage of the pipeline shown in Figure~\brains B shows the effect of using dynamic time warping to temporally align an example participant’s recall topic proportions matrix with the episode’s topic proportions matrix.  The rightmost matrix (stage 1) shows the correlation between the topic vectors for each unwarped recall timepoint (i.e., sliding text window) and each episode timepoint (TR).  The middle matrix (stage 2) shows the correlation between the topic vectors for each warped recall timepoint (i.e., TR) and each episode timepoint (TR).  In both matrices, the row (episode timepoint) matched to each column (recall timepoint) by the dynamic time warping algorithm is indicated in yellow.  We chose to visualize this step by correlating the example recall trajectory with the episode (rather than with itself, as in the other matrices) before versus after warping.  We felt this would help to illustrate how the diagonal ``straightens'' as a result of the non-linear ``stretching'' the algorithm performs to align the two timeseries.  An important feature of the warping algorithm is that it is a monotonic transformation-- i.e., it does not change the relative orders of the recalled timepoints; it only stretches or compresses different parts of the recall topic proportions matrix while preserving its temporal order.

The 3rd (leftmost) stage shown in Fig.~\brains B then displays the proximal correlation matrix for the example participant’s recalls.  In other words, we computed the correlation matrix for that participant’s warped recall topic proportions (i.e., after they have been temporally aligned to the episode), and then masked out everything beyond the $n$th diagonal.  We label axes corresponding to the warped recall trajectory as “Warped recall time (TR)” rather than “Viewing time (TR)” to help differentiate them from axes corresponding to the episode trajectory, as well as to maintain consistency with how we labeled video-recall and recall-recall correlation matrices in other figures.

@@

\begin{figure}[tp]
\centering
\includegraphics[width=0.6\textwidth]{figs/topic_space_flow}
\caption{\small \textbf{Methods detail for recall trajectory analysis displayed in Figure~\trajectories B} \textbf{A.} This panel replicates Figure~\trajectories B in the \textit{Main text}, but with two additions.  First, individual participants' recall trajectories are displayed (faintly) as light gray lines.  Second, three locations on the trajectory have been highlighted (blue, yellow, and red circles).  \textbf{B.}  These zoomed-in views of the locations highlighted in Panel A show the average trajectory (black) and individual participants' trajectories (gray lines) that intersect the given region of topic space.  \textbf{C.} For each circular region of topic space tiling the 2D embedding plane displayed in Panel A, we compute the distribution of angles formed between each participant's trajectory segment (i.e., the point at which the trajectory enters and exists the region of topic space) and the $x$-axis.  The distributions of angles for these three example regions are displayed in the colored rose plots.  We use Rayleigh tests to assign an arrow direction, length, and color for that region of topic space.  Arrows displayed in color are significant at the $p < 0.05$ level (corrected).  The arrow directions are oriented according to the circular means of each distribution, and the arrow lengths are proportional to the lengths of those mean vectors.  The example regions have been oriented from left to right in decreasing order of consistency across participants.}
\label{fig:arrows-methods}
\end{figure}
\vspace*{\fill}
